Analysis of CSV File Reading in R

### 2021-12-10 paper.Rmd
nite

The goal of this analysis is to determine the effects of three factors - `R` package, file size, computer device - on the reading time for CSV files measured in seconds. Because of this, we chose to design our experiment with a  $3^3$ factorial experiment [@montgomery] in mind. Furthermore, we chose to include $n=20$ replicates for each combination of our factors.  Our chosen design implies that the total number of observations for our experiment is 540. Our data was generated by considering the $3^3$ possible combinations of `R` package, file size, and computer. We then obtained the twenty replicates for each combination according to the order of a random sample of 1 through $3^3$.

The three levels for each of our factors are described as follows. For our first factor, R functions and their associated packages, we considered the three package-function combinations as introduced in Section 1. This includes  `rio::import()`, `data.table::fread()`, and `readr::read_csv()`. As for our second variable, file size, we utilized `R` to generate and export three different CSV files of different sizes. Each data file consists of columns of randomly-draw samples from a standard normal distribution. First, the largest file consists of 1000000 rows and 100 columns and has a size of 0.49 GB. Next, we have a mid-sized file of 1.22 GB on disk containing 1000000 cases and 62 variables. Finally, our smallest CSV file has a row-by-column dimension of 252000 by 100 and takes up 1.96 GB of storage. Last but not least, we considered three different laptop devices for this experiment. In particular, we have a MacBook Air 2020 with an M1 processor, a Lenovo Legion 2019 with a Ryzen 7 4800H processor, and a Dell XPS 13 2020 with an  Intel i7 processor.

@@

After gaining an understanding of our data, the next step was to fit the model corresponding to our experiment. The model for a three-level factorial experiment with three factors is given by

\begin{equation}
y_{ijkl} = \mu + \tau_i + \beta_j + \gamma_k + (\tau \beta)_{ij} + (\tau \gamma)_{ik} + (\beta \gamma)_{jk} + (\tau \beta \gamma)_{ijk} + \epsilon_{ijkl},
\end{equation}
where $i, j , k = 1, 2, 3$ and $l = 1, 2, ...,20$, since we have $n=20$ replicates.

@@

# Discussion

>>>>

The goal of this analysis is to determine the effects of three factors - `R` package, file size, computer device - on the reading time for CSV files measured in seconds. Because of this, we chose to design our experiment with a  $3^3$ factorial experiment [@montgomery] in mind. Furthermore, we chose to include $n=20$ replicates for each combination of our factors.  Our chosen design implies that the total number of observations for our experiment is 540.

The three levels for each of our factors are described as follows. For our first factor, R functions and their associated packages, we considered the three package-function combinations as introduced in Section 1. This includes  `rio::import()`, `data.table::fread()`, and `readr::read_csv()`. As for our second variable, file size, we first utilized `R` to generate and export three CSV files of different sizes consisting of columns of randomly-draw samples from a standard normal distribution. First, the largest file consists of 1000000 rows and 100 columns and has a size of 0.49 GB. Next, we have a mid-sized file of 1.22 GB on disk containing 1000000 cases and 62 variables. Finally, our smallest CSV file has a row-by-column dimension of 252000 by 100 and takes up 1.96 GB of storage. Last but not least, we considered three different laptop devices for this experiment. In particular, we have a MacBook Air 2020 with an M1 processor, a Lenovo Legion 2019 with a Ryzen 7 4800H processor, and a Dell XPS 13 2020 with an  Intel i7 processor.

Our data was generated by considering the $3^3$ possible combinations of `R` package, file size, and computer. For each combination of reading function, file size, and device, we utilized the `R` function `Sys.time()` to measure the time it took to load the CSV file into our statistical software. We obtained the twenty replicates for each combination according to the order of a random sample of 1 through $3^3$.

@@

After gaining an understanding of our data, the next step was to fit the model corresponding to our experiment. The model for a three-level factorial experiment with three factors is given by \begin{equation} y_{ijkl} = \mu + \tau_i + \beta_j + \gamma_k + (\tau \beta)_{ij} + (\tau \gamma)_{ik} + (\beta \gamma)_{jk} + (\tau \beta \gamma)_{ijk} + \epsilon_{ijkl}, \end{equation} where $i, j , k = 1, 2, 3$ and $l = 1, 2, ...,20$, since we have $n=20$ replicates.

@@

# Conclusion and Discussion

In this paper, we examined the effects of three factors file size, computer device, and importing function on the reading time for CSV files in the `R` statistical programming language. While we could not conduct an ANOVA on our three-level factorial experiment due to the normality departures, we were still able to find results through a nonparametric permutation test. The results of our permutation test for the corresponding factorial design indicate that at least one level for all of the factors in our experiment (`R` package/function, file size, and computer) significantly differ in average reading speed time. In particular, the `rio` package resulted in significantly different mean CSV reading times compared to the packages `readr` and `data.table`. As for laptop device, the Lenovo machine had significantly lower mean importing times compared to the MacBook and Dell laptops. Lastly, the average time elapsed for loading CSV files into `R` differ across the three file sizes of 0.49 GB, 1.22 GB, and 1.96 GB. An additional analysis using ANCOVA suggests that `readr` is the most time-efficient package for getting CSV files into R on a MacBook Air with an M1 processor.

In the wake of these results, we realize that there are numerous factors that we control for in future studies. While we ran this experiment with all three laptops on battery power, we hypothesize that there is a significant difference between file speeds for a plugged-in laptop versus a laptop on battery power. Another variable that may be significant is the size of memory within the user’s `R` environment. In addition, we could consider loading data in different environments, such as RStudio Desktop, RStudio Cloud, Google Colab Notebook, and Terminal/Command line. We could also look into the reading time for other forms of data files, such as `.tsv`, `.xlsx`, or `.json`. We also wonder if there is a significant difference in speed for techniques for reading other forms of data, such as image, text, or sound data. Future work may also include implementing a similar design with more than twenty replicates in order to improve the fit of the model and proceed with a parametric approach. Overall, this paper provides insight into the first step of the problem of working with big data- loading the data. Moving forward, more discussion between computer scientists and statisticians will allow us to more quickly fill computational pitfalls that arise when working with big data.
